---
title: |
  |  \Large Eras in baseball: Change point analysis fun title. 
author: | 
  | \large Mena Whalen \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`mwhalen3@luc.edu`](mailto:mwhalen3@luc.edu) \vspace{-1mm}
  |
  | \large Gregory J. Matthews \vspace{-1.1mm}
  | \normalsize Department of Mathematics and Statistics \vspace{-1mm}
  | \normalsize Loyola University Chicago \vspace{-1mm}
  | \normalsize Chicago, IL 60660 \vspace{-1mm}
  | \normalsize [`gmatthews1@luc.edu`](mailto:gmatthews1@luc.edu) \vspace{-1mm}
abstract: |
  | Baseball is some weird and wild shit. \vspace{2mm}
  | *Keywords*: change point analysis, baseball, 
bibliography: references.bib
fontsize: 12pt
link-citations: true
linkcolor: cyan
urlcolor: cyan
output:
  pdf_document:
    df_print: kable
    number_sections: true
    keep_tex: true
header-includes:
 \usepackage{setspace}
 \setstretch{1.15}
 \usepackage{float}
 \floatplacement{figure}{t}
---

```{r setup, echo=FALSE, message=FALSE}
knitr::opts_chunk$set(
	echo = FALSE,
	fig.align = "center",
	message = FALSE,
	warning = FALSE
)

gjm <- function(x, color = "red") {
  sprintf("\\textcolor{%s}{%s}", color, x)}
```

```{r pkgs}
library(tidyverse)
theme_set(theme_minimal())
library(Lahman)
```

\newpage

# Introduction {#sec:intro}
The first professional baseball team in the United States, the Cinicnnati Red Stockings, was formed in 1869.  Many leagues came and went in the late 1800s, but National League (NL), formed in 1876, emerged as the predominant league of the time.  A few decades later, the American League (AL) began growing in popularity and eventually reached an agreement with the NL to be the two major leagues of baseball with the winner of each league playing in the World Series starting in 1903.  

Throughout the history of baseball, the game has gone through many changes and eras.  For example, the time period between approximately 1901-1919 is often referred to as the "Dead Ball Era" and was characterized by low scoring games and dominant pitching.  Another more recent example would be the "Steroid Era" which lasted from approximately 1994 through 2005.  

SOME MORE STUFF

The goal of this work is to use multivariate change point analysis to identify change points in MLB as a whole to objectively identify eras in the games history, but also to look for change points to identify team specific eras.  

The work that is the most closely related to our own is that of @Groothius2017, which looked for structural breaks (Note: the terms "structural breaks" and "change points" are used interchangeably in this manuscript) in univariate times series.  

univariate time series of the means and SD's of several statistics (i.e. SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrane Multiplier (LM) unit root test @LeeandStrazicich2003.  Notably they find structural breaks in slugging percentage in 1921 and 1992.  In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).


From Woltring:
"Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras. A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005) (17). This study runs through the 2011 season and a seventh era will be added and labeled the Post Steroid Era (2006-2011)"

http://mlb.mlb.com/mlb/history/mlb_history_people.jsp?story=com



@Groothius2017 is the most closely related work to ours.  They use univariate time series of the means and SD's of several statistics (i.e. SLUG, HR rate, RBI rate, and batting average) to look for structural breaks via the Lagrane Multiplier (LM) unit root test @LeeandStrazicich2003.  Notably they find structural breaks in slugging percentage in 1921 and 1992.  In contrast to their methods, we use  more modern methods Double CUSUM Binary Segmentation algorithm (@Cho2016) and the Sparsified Binary Segmentation algorithm (@ChoFryzlewwicz2014).  







@Berry1999bridging they talk about bridging eras.  

When did the steroids era start: https://www.espn.com/mlb/topics/_/page/the-steroids-era#:~:text=Unlike%20other%20MLB%20%22eras%2C%22,leaguewide%20PED%20testing%20until%202003.

Traditional wisdom:
1900-1919 dead ball era

From Woltring:
"Baseball has endured much change over the course of its history, and because of constant change, the modern era of baseball has been segmented into six distinct sub-eras. A common list presented at Baseball-Reference described the eras as the Dead Ball Era (1901-1919), the Live Ball Era (1920-1941), the Integration Era (1942-1960), the Expansion Era (1961-1976), the Free Agency Era (1977-1993) and the Long Ball/Steroid Era (1994-2005) (17). This study runs through the 2011 season and a seventh era will be added and labeled the Post Steroid Era (2006-2011)"


https://www.baseball-reference.com/bullpen/Deadball_Era
1901-1920

Mound was lowered in december 1968.  
https://www.baseball-reference.com/bullpen/Pitcher%27s_mound

https://www.espn.com/mlb/story/_/id/33238595/major-league-baseball-stops-testing-players-steroids-nearly-20-years-report-says

https://www.semanticscholar.org/paper/Was-There-a-Structural-Break-in-Barry-Bonds's-Bat-Nieswiadomy-Strazicich/0a32effa10a3d26ea4675efcca3f13229efe0f7d


# Methods
@ChoFryzlewwicz2014 and @Cho2016

R Pacakge: @hdbinseg

# Results
```{r echo = FALSE}
library(tidyverse)
library(janitor)
library(hdbinseg)

teams <- read_csv("../data/Teams.csv") %>% 
  clean_names()


team_names <- teams %>% 
  group_by(franch_id) %>% 
  count() %>% 
  filter(n > 120) %>% 
  pull(franch_id)

# teams %>% 
#   filter(franch_id %in% team_names) %>% 
#   group_by(franch_id) %>% 
#   summarise(min(year_id)) ## filter after 1901

old_teams_full <- teams %>% 
  filter(franch_id %in% team_names & year_id > 1900) %>% 
  select(year_id, franch_id, w, l, r, ab, h, hr, bb, so)


teams_wide <- old_teams_full %>% 
  pivot_wider(id_cols = franch_id, 
              names_from = year_id, 
              values_from = hr)

## find thresholding value for cpt using bootstrapping
thrs <- dcbs.thr(as.matrix(teams_wide[-1]), interval = c(1, dim(teams_wide[-1])[2]))
### if I lower the threshold more changepoints
hr_cpt_dcbs <- dcbs.alg(as.matrix(teams_wide[-1]), cp.type = 1, thr = thrs-1)
hr_cpt_dcbs$ecp
hr_cpt_sbs <- sbs.alg(as.matrix(teams_wide[-1]), cp.type = 1)         
hr_cpt_sbs$ecp

cpt_years <- as.numeric(names(teams_wide[-1])[hr_cpt_dcbs$ecp])

ggplot(old_teams_full, aes(year_id, hr, color = franch_id)) +
  geom_line() +
  geom_vline(xintercept = cpt_years)


```



## Results

```{r}
load("/Users/gregorymatthews/Dropbox/baseball_cpt/data/teams_cpt_time_df.RData")
teams_cpt_time_df

```

NYY 1918: Babe Ruth got to the Yankees in 1920.  And starting in 1919 they had exactly one losing seasons between 1919 and 1965.  
BOS 1917, 1929, 1948, 1994: 
Their last world series win in the 1900s was in 1918.  
I don't know 1929. 
1948 there was a big jump in runs?  
1994: Strike year. 

LAD: 1919, 1955

ATL 1918 1949

CHW: 1919: Blaack Sox Scandal

CHC: 1919 1940
1940: War. 

Cin: 1929 1952

CLE 1920 1939
Cleveland won the world series in 1920. Major change in offensive output.  


DET 1927 1940

BAL 1919 1942

SFG 1919 1937

OAK 1920 1936 1951

PHI 1918 1937

PIT 1920 1946

STL 1919 1941

MIN 1919 1942

Pearl Harbor was 1941.  So US was in war in 1942.  




# Acknowledgements {-}


We thank Michael Lopez for suggesting we do "something with change point analysis".  

# Supplementary Material {-}

All code for reproducing the analyses in this paper is publicly available at https://github.com/menawhalen/baseball_cpt

# References
